---
title: "Week 10: Exploring DHS data with ggplot2"
output:
  prettydoc::html_pretty:
    theme: architect
    toc: true
author: "GRAPH Courses Team"
date: "2023-09-12"
editor_options: 
  chunk_output_type: inline
---

```{r echo = F}
knitr::opts_chunk$set(warning = F, message = F)
```

## Demographic and Health Survey (DHS) data

The Demographic and Health Surveys (DHS) are a series of standardized surveys conducted periodically in developing countries that collect data on various aspects of population, health, and nutrition. These surveys are a gold-standard source of reliable public health data for many countries and international bodies. You will be using DHS data for the final project of this course, so this is a chance to practice working with this type of data.

Several datasets (called recodes) are made available from each survey, including a household recode, men's recode and the women's recode (sometimes called the individual recode). This ***individual recode*** dataset contains individual-level data on women aged 15 to 49 in the surveyed households. This is what you will use for this workshop and for your final projects.

Today we will be looking at individual recode (IR) data from **Nigeria (2018)**. Key finding from this dataset can be found in this [report](https://drive.google.com/file/d/10pp9O9EJZvideynYyANqJtBCsN_vNQjs/view).

### Load packages and importing data

The dataset from DHS comes in the form of a **.DTA file** (the main data file in STATA format). If you look in your "data" folder of this workshop, you will see a file named NGIR7BFL_mini.DTA, which contains the IR data we want to use.

To import the DHS dataset from the .DTA format into R, you should use the `read_dta()` function from the {haven} package.

```{r}
# Load packages
pacman::p_load(
  janitor, # data analysis utilities
  here, # force rmds to use the project folder as working directory
  haven, # for reading in stata files
  tidyverse # for everything!
)
```

DHS files can be quite large -- the Nigeria 2018 data has over 40,000 rows and 5,000 columns.

In this demo, we created a mini version that has only 1000 rows. But in your actual final project, you will be expected to work with all rows.

When datasets are that large, you need to avoid importing the entire dataset at once, as this may take unreasonably long to run and may cause R to freeze. Instead, you can import only a subset of the data containing the variables you want to analyze. Consult the data dictionary (NGIR7BFL_data_dictionary.pdf) in the "data" folder to choose variables.

For this demo, we will select basic demographic variables (age, location, education, etc.) and a few additional variables related to fertility and reproductive timing among women in Nigeria.

Use the `col_select` argument of `read_dta()` to specify which columns to import.

```{r}
# Reading in the .dta file
ir_raw <- 
  haven::read_dta(here("data/NGIR7BFL_mini.DTA"),
                  col_select = c(
                    # basic information
                    v006, v007, v009, v010, v012, v013, 
                    v022, v024, v025, v106, v130, v190,   
                    # fertility and reproductive timing
                    v201, v212, v213, v511, v531))
```

Since the DHS variable names are not very descriptive (like v000 and v001), we will use `rename()` to give them short, descriptive titles.

```{r}
# Renaming the variables
ir_renamed <- 
  ir_raw %>% 
  rename(mth_interview = v006,
         yr_interview = v007,
         mth_birth = v009,
         yr_birth = v010,
         age = v012,
         age_group = v013,
         state = v022,
         region = v024,
         urban_rural = v025,
         highest_educ = v106,
         religion = v130,
         wealth_index = v190,
         num_kids = v201,
         age_first_birth = v212,
         curr_pregnant = v213,
         age_first_cohabitation = v511,
         age_first_sex = v531)
```

Another note on importing data: the **`read_dta()`** function reads in factor data as a special data type called **labelled** data.

```{r}
# Review data types
glimpse(ir_renamed)
```

Several variables are of class "`dbl+lbl`". This mean the columns display only numbers, and each number corresponds to a survey answer.

This is cryptic and can be problematic for later analysis, so we recommend converting these to regular R factors using the function `haven::as_factor()`. This will replace the numbers with the corresponding text label.

For example:

```{r}
# convert from labelled data to regular factors
ir_clean <- 
  ir_renamed %>% 
  haven::as_factor()


```

**NOTE**: There is another function called `as_factor()` from the {forcats} package, and one called `as.factor()` from {base} R. Remember to add "`haven::`" before `as_factor()` to make sure you're using the correct one for .dta files.

Now you can see that the variables that were previously "`dbl+lbl`" are now "`fct`" (factor) and the numbers are replaced with descriptive text.

```{r}
# Review data types
glimpse(ir_clean)
```

## Distributions with histograms, boxplots, and barplots

```{r}
# Histogram of age at first birth
 ggplot(data = ir_clean,
   mapping = aes(x = age_first_birth, fill = urban_rural)) +
   geom_histogram(bins = 25, 
     color = "white",
     position = "dodge") +
   theme_classic()
```

```{r}
# Boxplot of age at first birth

 ggplot(data = ir_clean,
   mapping = aes(x = age_first_birth, y = highest_educ)) +
   geom_boxplot() +
   theme_classic()
 

 ggplot(data = ir_clean,
   mapping = aes(x = age_first_birth, y = highest_educ, 
     fill = urban_rural)) +
   geom_boxplot() +
   theme_classic()
```


Like histograms, bar plots also visualize distributions, but for categorical variables.

```{r}

# stacked bar plot of education
ggplot(data = ir_clean,
  mapping = aes(
    x = highest_educ,
    fill = urban_rural)) +
  geom_bar() +
  theme_bw()

# percent stacked bar plot of education
ggplot(data = ir_clean,
  mapping = aes(
    x = highest_educ,
    fill = urban_rural)) +
  geom_bar(position = "fill") +
  theme_bw()
```
