---
title: "Week 10: Exploring DHS data with ggplot2"
output:
  prettydoc::html_pretty:
    theme: architect
    toc: true
author: "YOUR NAME HERE"
date: "2023-09-12"
editor_options: 
  chunk_output_type: inline
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r}
# Load necessary packages
pacman::p_load(
  janitor,    # for data analysis utilities
  here,       # to set the working directory
  haven,      # for reading Stata files
  tidyverse   # for data manipulation and visualization
)
```

# Data Importation

The given code chunk imports the same variables as presented in the demo. However, we want to augment this by adding five variables related to malaria-knowledge.

Scan through the data dictionary file and determine which five variables correspond to the following agreement statements:

-   Malaria can be fully cured by medicine
-   Malaria can lead to death
-   No worry about malaria due to easy treatment
-   Know people sick with malaria
-   Only weak children can die from malaria

Then, incorporate these variables into the data import code below.

```{r}
# Import the .dta file
ir_ng_18_raw <- 
  haven::read_dta(here("data/NGIR7BFL_mini.DTA"),
                  col_select = c(
                    # Basic information
                    v006, v007, v009, v010, v012, v013, 
                    v022, v024, v025, v106, v130, v190,                     
                    
                    # Literacy and media
                    v155, v156, v157, v158, v159, 
                    
                    # Fertility and reproductive timing
                    v201, v212, v213, v511, v531
                    
                    # Add the identified malaria knowledge variables here
                    
                    ))
```

**Note**: make sure ALL letters in the variable names are in lower case. If the data dictionary says "S1108BA" you should change the letters to lowercase and enter "s1108ba" in `col_select`.

# Variable Renaming

Assign descriptive names to the newly imported variables and then execute the code below.

```{r}
# Rename the variables
ir_ng_18_renamed <- 
  ir_ng_18_raw %>% 
  rename(age = v012,
         age_group = v013,
         state = v022,
         region = v024,
         urban_rural = v025,
         highest_educ = v106,
         religion = v130,
         literacy = v155,
         wealth_index = v190,
         num_kids = v201,
         age_first_birth = v212,
         age_first_cohabitation = v511,
         age_first_sex = v531 # Rename malaria knowledge variables
         
         
         )
```

# Data Transformation

With the suitable function from the `haven` package, transform all variables into regular R factors.

```{r}
# Add function to convert from labelled data to regular factors
ir_ng_18_clean <- 
  ir_ng_18_renamed 
  

```

# Comparing distributions

## Histograms and boxplots for continuous distributions

a.  Choose one of the continuous variables from the dataset: `age`, `num_kids`, `age_first_sex`, `age_first_cohabitation` or `age_first_birth.` Then, generate a simple histogram of this variable. Make sure to provide the necessary labels.

b.  Now, incorporate the `fill` aesthetic to color the bars of your histogram by levels of a categorical variable. Map one of the new malaria knowledge variables to fill - `aes(..., fill = MALARIA_VAR)`, and use `position` adjustments to create a side-by-side (dodged) histogram. This graph should show how the responses to this question differ at the various levels of the chosen continuous variable.

c. Visualize the same comparisons with a boxplot instead of a histogram.

## Barplots for discrete distributions 

a.  Select one of the categorical variables from the dataset. Proceed to generate a simple bar plot to display the distribution of this variable. 

b.  Next, add the `fill` aesthetic, `aes(fill = CATEGORICAL_VAR)`, with `position` adjustments to generate a percentage-stacked bar plot. This graph should illustrate how the answers to the malaria knowledge question change at the different levels of the chosen categorical variable.


# Bonus challenge

Pick one of the plots from above to customize further, then implement the below customizations.

## Facets 

Another way of comparing distributions by levels of a categorical variable is **faceting**. For example:

```{r} 
ggplot(data = mtcars, aes(x = mpg)) +
  geom_histogram() +
  facet_wrap(~cyl)
```

Implement something similar in your plots, faceting by a selected categorical variable.


## Customizing colors

Rather than just let ggplot2 pick your fill colors for you, you can implement a custom palette using `scale_fill_manual()` and the `values` argument. 

To generate a custom palette, go to https://mycolor.space and enter this hex code, which is the GRAPH Network's logo color: `#6DA5AF`. Then generate a suitable palette based on this color. 

Copy the hex codes and use them in the `scale_fill_manual()` function as done in the example below:

```{r}
# Histogram with custom colors
ggplot(mtcars, aes(x = mpg, fill = factor(cyl))) +
  geom_histogram(position = "dodge", binwidth = 2) +
  scale_fill_manual(values = c("#4C5B5C", "#6B8E7A", "#EFF2A0")) 
```

