---
title: "Workshop 6: Grouping, Summarizing and Plotting"
output:
  prettydoc::html_pretty:
    theme: architect
    toc: true
    number_sections: false
author: "Author's name here"
date: "2023-08-15"
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

```{r include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Load packages and data

To get started, **load in the needed packages**: {tidyverse}, {here},
{janitor}, and {esquisse}.

```{r}
pacman::p_load(tidyverse, esquisse, here, janitor)
```

Now, **read the dataset into R**. The data frame you import should have
880 rows and 22 columns. Remember to use the `here()` function to allow
your Rmd to use project-relative paths.

```{r}
# Load dataset
tb_raw <- read_csv(here::here("data/india_tuberculosis.csv"))
```

The column names in this CSV have spaces in them, which is not
R-friendly. I would recommend using `janitor::clean_names()` to give
your variable names a clean and consistent format.

```{r}
# Clean variable names
tb_renamed <- janitor::clean_names(tb_raw)
```

**Checkpoint**: The dataframe should contain 880 rows and 21 columns
(see Environment tab). Column/variable names should be all lowercase
with no spaces. You can run `names(tb_renamed)` to print the variable
names.

# Investigating healthcare costs

## Step 1: Calculate a `total_` variable

There were three visits for the participants and each had a cost
(`first_visit_cost, second_visit_cost, third_visit_cost`). Add together
these costs to create a `total_cost` variable.

```{r}
tb_clean1 <- tb_renamed %>%
  mutate(total_cost = first_visit_cost + second_visit_cost + third_visit_cost)
```

## Step 2: Summarize costs by group

Let's compare the cost of treatment at different health facilities using a summary table. Use `dplyr` verbs to group by `first_visit_location` and summarize the mean first visit cost.

```{r}
tb_clean1 %>% 
  group_by(first_visit_location) %>% 
  summarise(
    mean_first_cost = mean(first_visit_cost))
```

Reorder the rows of the summary table to go from highest to lowest mean cost. 

```{r}
cost_table <- tb_clean1 %>% 
  group_by(first_visit_location) %>% 
  summarise(
    mean_first_cost = mean(first_visit_cost)) %>% 
  arrange(desc(mean_first_cost), .by_group = TRUE)
```



# Encoding age groups

## Step 1: Create an `age_group` variable

The `age` variable records the age of each patient in years. For further
manipulations, we want to classify the patients into 4 equally-sized age
groups (i.e., the number of patients in each age group should be
approximately the same).

In order to determine what the age range for each age group should be,
we can use the `quantile()` function.

```{r}
quantile(tb_clean1$age)
```

We will now choose cutoffs for each age group based on these values.

**Create a new `age_group` variable with the following categories:**

-   18-36

-   37-47

-   48-57

-   58+

```{r}
tb_clean2 <- tb_clean1 %>% 
  mutate(
    age_group = case_when(
      age <= 36 ~ "18 to 36",
      age <= 47 ~ "37 to 47",
      age <= 57 ~ "48 to 57",
      age > 57 ~ "over 57"
    ))

# Note: we do not need to specify age ranges in each condition, because the first condition takes precedence over the subsequent conditions. However, using `&` to specify age ranges is also correct.
```

Now we can create a table of the `age_group` variable to see if we met
our goal of having a similar number of patients in each age group:

```{r}
# Tabulate age groups
tb_clean2 %>% tabyl(age_group)
```

**Checkpoint**: The if you classified the age groups correctly, you will
see that each age group has 24-36% of the patients.

## Step 2: Summarize costs by age group and smoking status

Use "nested" grouping to group the data by two variables: `age_group` and `smoking`. Then filter to get the most expensive `total_cost` for  each nested group. 

```{r}
tb_clean2 %>% 
  group_by(smoking, age_group) %>% 
  filter(total_cost == max(total_cost))
```

# Visualize data with {esquisse}


Using `esquisser()` and the costs summary table you created earlier, create a bar plot of mean costs, by treatment location.

```{r}
"PASTE_THE_ESQUISSE_CODE_HERE"

ggplot(cost_table) +
  aes(
    x = first_visit_location,
    y = mean_first_cost,
    fill = first_visit_location
  ) +
  geom_col() +
  scale_fill_hue(direction = 1) +
  theme_minimal()

```


# Wrap up

That's it for this assignment! We will choose 2-3 people to present your work during the workshop. If you would like to share your results with the class, please let an instructor know.

If you finish early and have extra time, you can explore the dataset further with esquisse, and create new plots to share with the class. Try customizing the colors and labels in your plots. 


